---
title: "Installing R and RStudio"
author: \textbf{Will Drysdale and Jack Davison}
institute: "University of York"
output: 
  beamer_presentation: 
    latex_engine: xelatex
    colortheme: dolphin
    theme: metropolis
urlcolor: blue
monofont: Consolas
header-includes:
- \newcommand{\blarge}{\begin{Large}}
- \newcommand{\elarge}{\end{Large}}
- \setmonofont[Colour=D40000]{Consolas}
---

```{r echo=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

## Purpose

- These slides guide you through the process of installing R and RStudio onto a desktop.

- R, RStudio and R Packages are three distinct things:
  - R is a **programming language** built for statistics.
  - RStudio is a **development environment** in which R runs. It makes R easier to use.
  - Packages are **collections of specialist tools** which make R more suited for certain tasks.

- Instructions are given for Windows computers, but should be the same for Mac and Linux; simply click on the appropriate option when they are presented.
  
## How to Install R

1.  Go to [cloud.r-project.org](https://cloud.r-project.org/).

2.  Select **Download R for Windows**.

3.  Select **base**.

4.  Click **Download R 4.1.2 for Windows**.

5.  **Run the .exe** to install R.

## How to Install RStudio

1.  Go to [rstudio.com](https://www.rstudio.com/).

2.  Navigate to **products** > **RStudio**.

3.  Scroll down and select **Download Rstudio Desktop**.

4.  Select **RStudio Desktop – Open Source License** (Free option).

5.  Select **RStudio - Windows**.

6.  **Run the .exe** to install RStudio.

## How to Install Packages

1.  **Launch RStudio**.

2.  In the console (the place you can type on the left-hand side of the screen, where the line starts with a “>”) type the following commands, pressing enter between them:

```{r chunk, eval = F}
install.packages("openair")
install.packages("tidyverse")
```

